Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2020-07-10 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Make layout (incl. newpage, etc)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3061 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 06f7f3b2848eb52e139661b2e7bf5746b0cc1cb5

\Abstract{
%SUBMITTED VERSION:
%While circular data occur in a wide range of scientific fields, the methodology
%for distributional modeling and probabilistic forecasting of circular response
%variables is rather limited. Most of the existing methods are built on the
%framework of generalized linear and additive models, which are often
%challenging to optimize and interpret. Therefore, building on previous ideas
%for trees modeling circular means, we suggest a distributional approach for
%regression trees and random forests yielding probabilistic forecasts based on
%the von~Mises distribution. The resulting tree-based models simplify the
%estimation process by using the available covariates for partitioning the data
%into sufficiently homogeneous subgroups so that a simple von~Mises distribution
%without further covariates can be fitted to the circular response in each
%subgroup. These circular regression trees are straightforward to interpret, can
%capture nonlinear effects and interactions, and automatically select the
%relevant covariates that are associated with either location and/or scale
%changes in the von~Mises distribution. Combining an ensemble of circular
%regression trees to a circular regression forest can regularize and smooth the
%covariate effects. The new methods are evaluated in a case study on
%probabilistic wind direction forecasting at two Austrian airports, considering
%other common approaches as a benchmark.
%
%NEW LISA AND MORITZ VERSION
%Circular data occur in a wide range of scientific fields, however the
%methodology for distributional modeling and probabilistic forecasting of
%circular response variables is rather limited. While most of the existing
%methods are built within the framework of generalized linear and additive
%models, which are often challenging to optimize and interpret, we propose a
%distributional approach for regression trees yielding probabilistic forecasts
%based on the von~Mises distribution. The resulting tree-based models use the
%available covariates for partitioning the data into sufficiently homogeneous
%subgroups, allowing for a simple circular distributional fit in each subgroup.
%These circular regression trees are straightforward to interpret, can capture
%nonlinear effects and interactions, and automatically select the relevant
%covariates that are associated with changes in any distribution parameter.
%Ensembles or forests of such circular trees can further improve the forecasts
%by regularizing and stabilizing the model. The newly developed methodology is
%illustrated in an application on short-term probabilistic wind direction
%forecasting for Innsbruck and Vienna Airport required to direct airplanes to a
%safe landing. The predictive skill of the novel approaches is benchmarked with
%an additive regression model plus a persistency and a climatological model,
%employing the circular continuous ranked probability score.
While circular data occur in a wide range of scientific fields, the methodology 
for distributional modeling and probabilistic forecasting of circular response
variables is rather limited. Most of the existing methods are built on
generalized linear and additive models, which are often challenging
to optimize and interpret. Specifically, capturing abrupt changes or
interactions is not straightforward but often relevant, e.g., for modeling wind
directions subject to different wind regimes. Additionally, automatic covariate
selection is desirable when many predictor variables are available, as often the
case in weather forecasting.

To address these challenges we suggest a general distributional approach using
regression trees and random forests to obtain probabilistic forecasts for
circular responses. Using trees simplifies model estimation as
covariates are used only for partitioning the data and subsequently just a
simple von~Mises distribution is fitted in the resulting subgroups.
Circular regression trees are straightforward to interpret, can capture
nonlinear effects and interactions, and automatically select covariates
affecting location and/or scale in the von~Mises distribution. Circular random
forests regularize and smooth the effects from an ensemble of trees.

The new methods are applied to probabilistic wind direction forecasting at two
Austrian airports, considering other common approaches as a benchmark.}

@@

\subsection{Data}
%%
\begin{figure}[t]
  \setkeys{Gin}{width=1.0\textwidth}
  \centering
  \includegraphics{_plot_circforest_stationmap_combined.pdf}
  \caption{Overview of the study area for Innsbruck Airport (left) and Vienna
    International Airport (right). For Innsbruck, $4$~stations at the airport and
    $6$~stations along the intersecting valleys are used, whereas, for Vienna,
    $9$~stations at the airport and $13$~stations within the vicinity of
    approximately 30\,km are used. Elevation data are obtained from the TandDEM-X
    digital elevation model with a horizontal resolution of $90$~m
    \citep{Wessel:2018}.}
  \label{fig:overview} 
\end{figure}
%%
\label{sec:wind:data}
The circular response variable considered in this application is a
$10$\,min-average of wind direction measurements at Innsbruck Airport (INN) and
Vienna International Airport (VIE) on an hourly temporal resolution. Temporal
information and $1$-hourly resolved $10$\,min mean observations of various
meteorological quantities are used as predictor variables, including wind
direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
their vicinities. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within its vicinity of approximately 30\,km are used.
Figure~\ref{fig:overview} provides a topographical overview of the airports and
their surrounding areas with the station sites employed in the application. In
addition, we use derived quantities such as 3-hourly means, minima and maxima,
as well as 1-and 3-hourly temporal changes and spatial differences towards the
airport of the respective quantities. An overview of the employed data sets can
be found in Table~\ref{tab:data}.

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Extending the toolbox for modeling circular data, circular regression trees and
forests are established by coupling model-based recursive partitioning with the
von~Mises distribution. By separating the data into more homogeneous subgroups,
possible difficulties in circular regression are bypassed as covariates are
solely considered for splitting and group-specific models are fitted without
further covariates. In addition, by specifying the von~Mises distribution for
each node and allowing for splits in both distribution parameters $\mu$ and
$\kappa$, fully probabilistic forecasts are provided.

>>>>

\Abstract{
%SUBMITTED VERSION:
%While circular data occur in a wide range of scientific fields, the methodology
%for distributional modeling and probabilistic forecasting of circular response
%variables is rather limited. Most of the existing methods are built on the
%framework of generalized linear and additive models, which are often
%challenging to optimize and interpret. Therefore, building on previous ideas
%for trees modeling circular means, we suggest a distributional approach for
%regression trees and random forests yielding probabilistic forecasts based on
%the von~Mises distribution. The resulting tree-based models simplify the
%estimation process by using the available covariates for partitioning the data
%into sufficiently homogeneous subgroups so that a simple von~Mises distribution
%without further covariates can be fitted to the circular response in each
%subgroup. These circular regression trees are straightforward to interpret, can
%capture nonlinear effects and interactions, and automatically select the
%relevant covariates that are associated with either location and/or scale
%changes in the von~Mises distribution. Combining an ensemble of circular
%regression trees to a circular regression forest can regularize and smooth the
%covariate effects. The new methods are evaluated in a case study on
%probabilistic wind direction forecasting at two Austrian airports, considering
%other common approaches as a benchmark.
%
%NEW LISA AND MORITZ VERSION
%Circular data occur in a wide range of scientific fields, however the
%methodology for distributional modeling and probabilistic forecasting of
%circular response variables is rather limited. While most of the existing
%methods are built within the framework of generalized linear and additive
%models, which are often challenging to optimize and interpret, we propose a
%distributional approach for regression trees yielding probabilistic forecasts
%based on the von~Mises distribution. The resulting tree-based models use the
%available covariates for partitioning the data into sufficiently homogeneous
%subgroups, allowing for a simple circular distributional fit in each subgroup.
%These circular regression trees are straightforward to interpret, can capture
%nonlinear effects and interactions, and automatically select the relevant
%covariates that are associated with changes in any distribution parameter.
%Ensembles or forests of such circular trees can further improve the forecasts
%by regularizing and stabilizing the model. The newly developed methodology is
%illustrated in an application on short-term probabilistic wind direction
%forecasting for Innsbruck and Vienna Airport required to direct airplanes to a
%safe landing. The predictive skill of the novel approaches is benchmarked with
%an additive regression model plus a persistency and a climatological model,
%employing the circular continuous ranked probability score.
While circular data occur in a wide range of scientific fields, the methodology 
for distributional modeling and probabilistic forecasting of circular response
variables is rather limited. Most of the existing methods are built on
generalized linear and additive models, which are often challenging
to optimize and interpret. Specifically, capturing abrupt changes or
interactions is not straightforward but often relevant, e.g., for modeling wind
directions subject to different wind regimes. Additionally, automatic covariate
selection is desirable when many predictor variables are available, as often the
case in weather forecasting.
% 
To address these challenges we suggest a general distributional approach using
regression trees and random forests to obtain probabilistic forecasts for
circular responses. Using trees simplifies model estimation as
covariates are used only for partitioning the data and subsequently just a
simple von~Mises distribution is fitted in the resulting subgroups.
Circular regression trees are straightforward to interpret, can capture
nonlinear effects and interactions, and automatically select covariates
affecting location and/or scale in the von~Mises distribution. Circular random
forests regularize and smooth the effects from an ensemble of trees.
%
The new methods are applied to probabilistic wind direction forecasting at two
Austrian airports, considering other common approaches as a benchmark.}

@@

\newpage
\subsection{Data}
%%
\begin{figure}[t]
  \setkeys{Gin}{width=1.0\textwidth}
  \centering
  \includegraphics{_plot_circforest_stationmap_combined.pdf}
  \caption{Overview of the study area for Innsbruck Airport (left) and Vienna
    International Airport (right). For Innsbruck, $4$~stations at the airport and
    $6$~stations along the intersecting valleys are used, whereas, for Vienna,
    $9$~stations at the airport and $13$~stations within the vicinity of
    approximately 30\,km are used. Elevation data are obtained from the TandDEM-X
    digital elevation model with a horizontal resolution of $90$~m
    \citep{Wessel:2018}.}
  \label{fig:overview} 
\end{figure}
%%
\label{sec:wind:data}
The circular response variable considered in this application is a
$10$\,min-average of wind direction measurements at Innsbruck Airport (INN) and
Vienna International Airport (VIE) on an hourly temporal resolution. Temporal
information and $1$-hourly resolved $10$\,min mean observations of various
meteorological quantities are used as predictor variables, including wind
direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
their vicinities. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within its vicinity of approximately 30\,km are used.
Figure~\ref{fig:overview} provides a topographical overview of the airports and
their surrounding areas with the station sites employed in the application. In
addition, we use derived quantities such as 3-hourly means, minima and maxima,
as well as 1-and 3-hourly temporal changes and spatial differences towards the
airport of the respective quantities. An overview of the employed data sets can
be found in Table~\ref{tab:data}.

@@

\newpage
% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Extending the toolbox for modeling circular data, circular regression trees and
forests are established by coupling model-based recursive partitioning with the
von~Mises distribution. By separating the data into more homogeneous subgroups,
possible difficulties in circular regression are bypassed as covariates are
solely considered for splitting and group-specific models are fitted without
further covariates. In addition, by specifying the von~Mises distribution for
each node and allowing for splits in both distribution parameters $\mu$ and
$\kappa$, fully probabilistic forecasts are provided.
